Automated Residual Plot Assessment with the R Package autovi and the Shiny App autovi.web

### 2024-11-28 autovi_paper.qmd
tweaked new paragraph

<!-- However, as discussed in @li2024plot, relying solely on subjective assessments of these plots can lead to issues, such as over-interpreting random patterns as model violations. -->

<!-- To diagnose the fit of a model it is recommended that the residuals are plotted. If the fit is good, any variation remaining should be noise, consistent with sampling from a distribution specified by the error model. Deviations that might be observed from a residual plot are non-normality, heteroscedasticity, and other associations with the fitted values.-->

@@

However, as discussed in @li2024automated, the lineup protocol has significant limitations in large-scale applications due to dependence on human labor. Thus, a computer vision model was developed with an associated statistical testing procedure to automate the assessment of residual plots. This model takes a residual plot and a vector of auxiliary variables (such as the number of observations) as inputs and outputs the predicted visual signal strength (VSS). This strength estimates the distance between the residual distribution of the fitted regression model and the reference distribution assumed under correct model specification.
<!---->
<!--Software for regression analysis tools is widely available. The Comprehensive R Archive Network (CRAN) [@hornik2012comprehensive] hosts a vast array of packages, many of which provide tools to diagnose models using residual plots. These packages can be broadly categorized into three groups: general-purpose, enhanced visual diagnostics, and visual diagnostics with statistical testing.-->
<!---->
<!--General-purpose regression analysis tools are the most widely used group of packages. While not specifically designed for graphical diagnostics of residuals in linear regression, they include this functionality as part of a broader suite of statistical tools. For example, the `stats` package [@stats] offers standard diagnostic plots such as residuals vs. fitted values, quantile-quantile (Q-Q) plots, and residuals vs. leverage plots. Additional packages like `jtools` [@jtools], `olsrr` [@olsrr], `rockchalk` [@rockchalk], and `ggResidpanel` [@ggresidpanel] provide similar graphical diagnostics, often with alternative aesthetics or interactive features. All of these tools deliver the types of diagnostic plots outlined in the classical text by @cook1982residuals. However, as discussed in @li2024plot, relying solely on subjective assessments of these plots can lead to issues, such as over-interpreting random patterns as model violations.-->
<!---->
<!-- The second group consists of enhanced visual diagnostics tools, which provide advanced aids for interpreting diagnostic plots. For instance, `ecostats` [@warton_global_2023] incorporates simulation envelopes into residual plots, while `DHARMa` [@dharma] compares empirical quantiles (0.25, 0.5, and 0.75) of scaled residuals to their theoretical counterparts. `DHARMa` is particularly focused on detecting model violations such as heteroscedasticity, incorrect functional forms, and issues specific to generalized linear and mixed-effect models, like over/under-dispersion. It also includes annotations from conventional tests, displayed as labels or text within the plot, to help avoid misinterpretation.-->
<!---->
<!-- The third group, visual diagnostics with statistical testing, offers tools for formally testing patterns observed in diagnostic plots [@buja2009statistical]. A common method for visual testing of residual plots is the lineup protocol, where the true residual plot is embedded within a set of null plots generated by simulating residuals under the null hypothesis $H_0$ that the regression model is correctly specified. Observers are asked to identify the plot that appears most different from the others. If a significant proportion of observers correctly identify the true residual plot, it provides evidence against $H_0$, as the true residual plot should be indistinguishable from the null plots if all residuals are generated by the same process, according to @buja2009statistical. Packages in this group, such as `nullabor` [@nullabor], `HLMdiag` [@loy2014hlmdiag], and `regressinator` [@regressinator], enable users to compare observed residual plots with samples from null distributions, helping to quantify the significance of any detected patterns. -->

<!-- However, as discussed in @li2024automated, the lineup protocol has significant limitations in large-scale applications due dependence on human labor. Thus a computer vision model was developed with an associated statistical testing procedure to automate the assessment of residual plots. This model takes a residual plot and a vector of auxiliary variables (such as the number of observations) as inputs and outputs the predicted visual signal strength (VSS). This strength estimates the distance between the residual distribution of the fitted regression model and the reference distribution assumed under correct model specification.-->
To make the statistical testing procedure and trained computer vision model widely accessible, we developed the R package `autovi`, and a web interface, `autovi.web` to make it easy for users to automatically read their residual plots with the trained computer vision model.

@@

Equation \ref{eq:kl-1} can be adapted to different forms to account for various model violations. For instance, in the case of a linear regression model where necessary higher-order predictors, $\boldsymbol{Z}$, and their corresponding parameter, $\boldsymbol{\beta}_Z$, are omitted, the distance measure can be expanded as follows:

@@

This distance measure depends on knowledge of the true residual distribution, which is often unknown. However, it can be estimated by identifying or approximating a mapping from a set of residuals to its corresponding distance measure. The computer vision model presented in @li2024automated approximates this mapping. It is trained on a large number of synthetic regression models, where the data-generating process is known, allowing the distance measure to be explicitly calculated. The model takes a residual plot as input and outputs the corresponding distance measure. Additional details are provided in @li2024automated.

@@

The main purpose of `autovi` is to provide rejection decisions and $p$-values for testing the null hypothesis ($H_0$) that the regression model is correctly specified. The package provides automated interpretation of residual plots using computer vision. The name `autovi` stands for **auto**mated **v**isual **i**nference. <!--While initially designed for linear regression residual diagnostics, it has the potential to be extended to broader visual inference applications, as we discuss in section @sec-autovi-infrastructure.-->
This functionality can be accessed through the R package `autovi`, or through a web interface, `autovi.web`, which allows use without the full installation of R, Python, and package dependencies on the user's system.  locally.

@@

The summary includes observed VSS of the true residual plot and associated $p$-value of the automated visual test. The $p$-value is the proportion of null plots (out of the total 100) that have VSS greater than or equal to that of the true residual plot. <!--We typically reject the null hypothesis when the $p$-value is smaller than or equal to a desired significance level, such as 5%.--> The report also provides sample quantiles of VSS for null samples and bootstrapped data plots, providing more information about the sampling variability and a likelihood of model violations. The likelihood is computed from the proportion of values greater than the observed VSS in both the bootstrapped data values and the simulated null values.

>>>>

@@

However, as discussed in @li2024automated, the lineup protocol has significant limitations in large-scale applications due to dependence on human labor. Thus, a computer vision model was developed with an associated statistical testing procedure to automate the assessment of residual plots. This model takes a residual plot and a vector of auxiliary variables (such as the number of observations) as inputs and outputs the predicted visual signal strength (VSS). This strength estimates the distance between the residual distribution of the fitted regression model and the reference distribution assumed under correct model specification.

To make the statistical testing procedure and trained computer vision model widely accessible, we developed the R package `autovi`, and a web interface, `autovi.web` to make it easy for users to automatically read their residual plots with the trained computer vision model.

@@

This distance measure depends on knowledge of the true residual distribution, which is unknown in practice. To compute $D_{KL}$ for the training samples, Equation \ref{eq:kl-1} takes different forms depending on the specific model violations. For instance, where necessary higher-order predictors, $\boldsymbol{Z}$, and their corresponding parameter, $\boldsymbol{\beta}_Z$, are omitted from the fitted linear model, the distance measure can be expanded as follows:

@@

The computer vision model approximates this mapping from a set of residuals to its corresponding distance measure. It is trained on a large number of synthetic regression models, where the data-generating process is known, allowing the distance measure to be explicitly calculated. The model takes a residual plot as input and outputs the corresponding distance measure. Additional details are provided in @li2024automated.

@@

The main purpose of `autovi` is to provide rejection decisions and $p$-values for testing the null hypothesis ($H_0$) that the regression model is correctly specified. The package provides automated interpretation of residual plots using computer vision. The name `autovi` stands for **auto**mated **v**isual **i**nference. This functionality can be accessed through the R package `autovi`, or through a web interface, `autovi.web`, which allows use without the full installation of R, Python, and package dependencies on the user's system.  locally.

@@

The summary includes observed VSS of the true residual plot and associated $p$-value of the automated visual test. The $p$-value is the proportion of null plots (out of the total 100) that have VSS greater than or equal to that of the true residual plot. The report also provides sample quantiles of VSS for null samples and bootstrapped data plots, providing more information about the sampling variability and a likelihood of model violations. The likelihood is computed from the proportion of values greater than the observed VSS in both the bootstrapped data values and the simulated null values.

### 2024-11-28 autovi_paper.tex
tweaked new paragraph

\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{anzsj}
\usepackage{bookmark}

@@

Equation \ref{eq:kl-1} can be adapted to different forms to account for
various model violations. For instance, in the case of a linear
regression model where necessary higher-order predictors,
\(\boldsymbol{Z}\), and their corresponding parameter,
\(\boldsymbol{\beta}_Z\), are omitted, the distance measure can be
expanded as follows:

@@

This distance measure depends on knowledge of the true residual
distribution, which is often unknown. However, it can be estimated by
identifying or approximating a mapping from a set of residuals to its
corresponding distance measure. The computer vision model presented in
\citet{li2024automated} approximates this mapping. It is trained on a
large number of synthetic regression models, where the data-generating
process is known, allowing the distance measure to be explicitly
calculated. The model takes a residual plot as input and outputs the
corresponding distance measure. Additional details are provided in
\citet{li2024automated}.

>>>>

\usepackage[]{natbib}
\bibliographystyle{anzsj}
\usepackage{bookmark}

@@

This distance measure depends on knowledge of the true residual
distribution, which is unknown in practice. To compute \(D_{KL}\) for
the training samples, Equation \ref{eq:kl-1} takes different forms
depending on the specific model violations. For instance, where
necessary higher-order predictors, \(\boldsymbol{Z}\), and their
corresponding parameter, \(\boldsymbol{\beta}_Z\), are omitted from the
fitted linear model, the distance measure can be expanded as follows:

@@

The computer vision model approximates this mapping from a set of
residuals to its corresponding distance measure. It is trained on a
large number of synthetic regression models, where the data-generating
process is known, allowing the distance measure to be explicitly
calculated. The model takes a residual plot as input and outputs the
corresponding distance measure. Additional details are provided in
\citet{li2024automated}.
